F\# Data: \textnormal{Making structured data first-class citizens}

### 2015-02-27 icfp-2015/paper.tex
Adding conclusion and more refs

\paragraph{Type providers.}
The notation $\ident{JsonProvider}\langl\str{people.json}\rangl$ on the first line passes a 
\emph{static parameter} to the type provider. Static parameters are resolved at compile-time, 
so the file name has to be a constant. The provider analyzes the sample and generates
a type that we name \ident{People}. In F\# editors that use the F\# Compiler Service [?], the 
type provider is also executed at development-time and so the same provided types are
used in code completion.

@@

\paragraph{Erasing type providers.}
At runtime, the type providers we describe use an erasure mechanism similar to Java Generics [?]. 
A type provider also generates code that is executed in place of the members of the generated types.
In the above example, the compiled (and actually executed) code looks as follows:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{air}~=~\ident{CsvFile}\ident{.Load}(\str{airdata.csv}, \kvd{fun}~\ident{r}\rightarrow\\[0.1em]
 \quad \ident{asDecimal r}.[\num{0}], \ident{asIntOpt r}.[\num{1}], \ident{r}.[\num{2}], \ident{asBool r}.[\num{3}])\\[0.6em]
 \kvd{for}~\ident{c1},\_,\ident{c3},\ident{c4} ~\kvd{in}~\ident{air.Rows}~\kvd{do}\\[0.1em]
 \quad\kvd{if}~\ident{not}~\ident{c4}~\kvd{then}~\ident{printf}~\str{\%s: \%d}~\ident{c3 c1}\\
\end{array}
\end{equation*}
%
The generated type \ident{AirCsv} is erased to a type $\ident{CsvFile}\langl\alpha\rangl$.
The \ident{Load} method takes the file name together with a function that turns a row represented 
as an array of strings to a typed representation -- here, a four-element tuple 
$\ident{decimal}*\ident{option}\langl\ident{int}\rangl*\ident{string}*\ident{bool}$ (this type
is also used as a type argument of $\ident{CsvFile}\langl\alpha\rangl$. The properties
such as \ident{Ozone} are then replaced with an accessor that reads the corresponding tuple
element.

The type erasure mechanism is explained in our simplified formal model in Section~\ref{sec:formal}.
More details about the full implementation in F\# can be found in the paper on F\# type 
providers [?].

@@

From looking at just the previous three examples, these may appear arbitrary, but our experience
suggests that they are the most common issues. The following JSON response with government debt
information returned by the World Bank API [?] demonstrates all three problems:
%
{\small{
\begin{verbatim}
  [ { "page": 1, "pages": 5 },
    [ { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2012",
        "value": null },
      { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2010",
        "value": "35.1422970266502" } ] ]
\end{verbatim}
}}
%
\noindent
First of all, the top-level element is a collection containing two values of different kind.
The first is a record with meta-data about the current page and the second is an array with data. 
The JSON type provider infers this as heterogeneous collection with properties \ident{Record}
for the former and \ident{Array} for the latter. Second, the \ident{value} is \kvd{null} for some
records. Third, numbers can be represented in JSON as numeric literals (without quotes), but
here, they are returned as string literals instead\footnote{This is often used to avoid non-standard
numerical types in JavaScript.}.

@@

We focus on the object system (which is what F\# type providers produce) and so the following combines 
standard ML [?] features with Featherweight Java [?] classes. However, we only need classes with properties and
without inheritance. A class has a single implicit constructor and the declaration closes over 
constructor parameters. To avoid including all of ML, we only select constructs for working with options and lists
that we need later.
%
\begin{equation*}
\begin{array}{rcl}
 \tau &\narrow{=}& \ident{int} \lsep \ident{decimal} \lsep \ident{float} \lsep \ident{bool} \lsep \ident{string} \\[0.0em]
      &\narrow{|}& C \lsep \ident{StructVal} \lsep \ident{list}\langl\tau\rangl \lsep \ident{option}\langl\tau\rangl \\[0.6em]
 L &\narrow{=}& \kvd{type}~C(\overline{x:\tau}) = \overline{M} \\[0.0em]
 M &\narrow{=}& \kvd{member}~N:\tau=e \\[0.6em]
 v &\narrow{=}& s \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \lsep [v_1; \ldots; v_n] \\[0.0em]
 e &\narrow{=}& op \lsep e.N \lsep \kvd{new}~C(\overline{e}) \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3}\\
   &\narrow{|}& \ident{Some}(e)\lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
   &\narrow{|}& \kvd{match}~e~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 \\
   &\narrow{|}& [e_1;\ldots;e_n]\lsep \ident{List.map}~(\lambda x\rightarrow e_1)~e_2
\end{array}
\end{equation*}

@@

The type providers for XML, CSV and JSON all work in the same way. They take sample documents and infer a common
supertype $\sigma$ from the samples. From $\sigma$, they then generate F\# types that are then exposed to the
programmer\footnote{The actual implementation provides \emph{erased types} as described in [?]. Here, we treat
the code as actually generated. This is an acceptable simplification, because F\# Data type providers do not 
rely on laziness that is available through erased types.}.

@@

Using the standard approach to syntactic type safety [?], we prove the type preservation 
(reduction does not change type) and progress (an expression that is not a value can be reduced).

@@

The \emph{relativized safety} property does not guarantee the same amount of safety as standard type safety
for programming langauges without type providers. However, it reflects the reality of programming with external
data sources that is increasingly important in the age of web [?]. So, type providers do not reduce the safety
-- they simply make the existing issues visible.

@@

Finally, the formal model presented here ignores two aspects of the real type provider mechanisms. As discussed
here, we do not use the \emph{type erasure} mechanism, but treat type providers as if they were actually 
generating code. For XML, CSV and JSON, both models can be used. However, one interesting aspect of erasing 
type providers is that the types can be generated lazilly -- only classes that are actually used in the type
checked code are generated. Extending our formalism to capture this aspect is an interesting future work
relevant to other type providers -- including the World Bank type provider [?] that is also available
in the F\# Data library.

@@

  - Prime example of type providers, explores relativized type safety
  - Used by industry, downloads, tested, etc. 
  - Predictable yet powerful type inference

\acks
Acknowledgments, if needed.

\cite{age-of-web}

>>>>

\paragraph{Type providers.}
The notation $\ident{JsonProvider}\langl\str{people.json}\rangl$ on the first line passes a 
\emph{static parameter} to the type provider. Static parameters are resolved at compile-time, 
so the file name has to be a constant. The provider analyzes the sample and generates
a type that we name \ident{People}. In F\# editors that use the F\# Compiler Service, the 
type provider is also executed at development-time and so the same provided types are
used in code completion.

@@

\paragraph{Erasing type providers.}
At runtime, the type providers we describe use an erasure mechanism similar to Java Generics \cite{java-erasure}. 
A type provider also generates code that is executed in place of the members of the generated types.
In the above example, the compiled (and actually executed) code looks as follows:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{air}~=~\ident{CsvFile}\ident{.Load}(\str{airdata.csv}, \kvd{fun}~\ident{r}\rightarrow\\[0.1em]
 \quad \ident{asDecimal r}.[\num{0}], \ident{asIntOpt r}.[\num{1}], \ident{r}.[\num{2}], \ident{asBool r}.[\num{3}])\\[0.6em]
 \kvd{for}~\ident{c1},\_,\ident{c3},\ident{c4} ~\kvd{in}~\ident{air.Rows}~\kvd{do}\\[0.1em]
 \quad\kvd{if}~\ident{not}~\ident{c4}~\kvd{then}~\ident{printf}~\str{\%s: \%d}~\ident{c3 c1}\\
\end{array}
\end{equation*}
%
The generated type \ident{AirCsv} is erased to a type $\ident{CsvFile}\langl\alpha\rangl$.
The \ident{Load} method takes the file name together with a function that turns a row represented 
as an array of strings to a typed representation -- here, a four-element tuple 
$\ident{decimal}*\ident{option}\langl\ident{int}\rangl*\ident{string}*\ident{bool}$ (this type
is also used as a type argument of $\ident{CsvFile}\langl\alpha\rangl$. The properties
such as \ident{Ozone} are then replaced with an accessor that reads the corresponding tuple
element.

The type erasure mechanism is explained in our simplified formal model in Section~\ref{sec:formal}.
More details about the full implementation in F\# can be found in the paper on F\# type 
providers \cite{fsharp-typeprov}.

@@

From looking at just the previous three examples, these may appear arbitrary, but our experience
suggests that they are the most common issues. The following JSON response with government debt
information returned by the World Bank API demonstrates all three problems:
%
{\small{
\begin{verbatim}
  [ { "page": 1, "pages": 5 },
    [ { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2012",
        "value": null },
      { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2010",
        "value": "35.1422970266502" } ] ]
\end{verbatim}
}}
%
\noindent
First of all, the top-level element is a collection containing two values of different kind.
The first is a record with meta-data about the current page and the second is an array with data. 
The JSON type provider infers this as heterogeneous collection with properties \ident{Record}
for the former and \ident{Array} for the latter. Second, the \ident{value} is \kvd{null} for some
records. Third, numbers can be represented in JSON as numeric literals (without quotes), but
here, they are returned as string literals instead\footnote{This is often used to avoid non-standard
numerical types in JavaScript.}.

@@

We focus on the object system (which is what F\# type providers produce) and so the following combines 
standard ML \cite{sml} features with Featherweight Java \cite{fwjava} classes. However, we only need classes with properties and
without inheritance. A class has a single implicit constructor and the declaration closes over 
constructor parameters. To avoid including all of ML, we only select constructs for working with options and lists
that we need later.
%
\begin{equation*}
\begin{array}{rcl}
 \tau &\narrow{=}& \ident{int} \lsep \ident{decimal} \lsep \ident{float} \lsep \ident{bool} \lsep \ident{string} \\[0.0em]
      &\narrow{|}& C \lsep \ident{StructVal} \lsep \ident{list}\langl\tau\rangl \lsep \ident{option}\langl\tau\rangl \\[0.6em]
 L &\narrow{=}& \kvd{type}~C(\overline{x:\tau}) = \overline{M} \\[0.0em]
 M &\narrow{=}& \kvd{member}~N:\tau=e \\[0.6em]
 v &\narrow{=}& s \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \lsep [v_1; \ldots; v_n] \\[0.0em]
 e &\narrow{=}& op \lsep e.N \lsep \kvd{new}~C(\overline{e}) \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3}\\
   &\narrow{|}& \ident{Some}(e)\lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
   &\narrow{|}& \kvd{match}~e~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 \\
   &\narrow{|}& [e_1;\ldots;e_n]\lsep \ident{List.map}~(\lambda x\rightarrow e_1)~e_2
\end{array}
\end{equation*}

@@

The type providers for XML, CSV and JSON all work in the same way. They take sample documents and infer a common
supertype $\sigma$ from the samples. From $\sigma$, they then generate F\# types that are then exposed to the
programmer\footnote{The actual implementation provides \emph{erased types} as described in \cite{fsharp-typeprov}. 
Here, we treat the code as actually generated. This is an acceptable simplification, because F\# Data type providers do not 
rely on laziness that is available through erased types.}.

@@

Using the standard approach to syntactic type safety  \cite{syntactic}, we prove the type preservation 
(reduction does not change type) and progress (an expression that is not a value can be reduced).

@@

The \emph{relativized safety} property does not guarantee the same amount of safety as standard type safety
for programming langauges without type providers. However, it reflects the reality of programming with external
data sources that is increasingly important in the age of web \cite{age-of-web}. So, type providers do not
reduce the safety -- they simply make the existing issues visible.

@@

Finally, the formal model presented here ignores two aspects of the real type provider mechanisms. As discussed
here, we do not use the \emph{type erasure} mechanism, but treat type providers as if they were actually 
generating code. For XML, CSV and JSON, both models can be used. However, one interesting aspect of erasing 
type providers is that the types can be generated lazilly -- only classes that are actually used in the type
checked code are generated. Extending our formalism to capture this aspect is an interesting future work
relevant to other type providers -- including the World Bank type provider that is also available
in the F\# Data library.

@@

In this paper, we explored the F\# Data type providers for structured data formats such as XML, CSV and JSON.
As most real-world data do not come with an explicit schema, the library uses \emph{type inference} that
deduces a type from a set of samples. Next, it uses the \emph{type provider} mechanism to integrate the 
inferred type directly into the F\# type system.

The type inference algorithm we use is based on a common supertype relation. For usability reasons, the
algorithm attempts to infer type that does not contain unions and prefers records with optional fields instead.
The algorithm is simple and predictable, which is important as developers need to understand how changing the
samples affects the resulting types.

In the second part of the paper, we explore the programming language theory behind type providers. F\# Data
is a prime example of type providers, but our work also demonstrates a more general point. The types generated
by type providers can depend on external input (such as samples) and so we can only formulate \emph{relativized
safety property}, which says that a program is safe only if the actual inputs satisfy additional conditions --
in our case, they have to be subtypes of one of the samples.

The type provider mechanism has been described before, but this paper is novel in that it explores concrete
type providers from the perspective of programming language theory. This is particularly important as the
F\# Data library is becoming de-facto standard tool for data access in F\#\footnote{At the time of writing, 
the library has 37,000 downloads on NuGet; over 1600 commits and 36 contributors on GitHub.} and other 
languages are begining to adopt mechanisms similar to F\# type providers.

\acks
We would like to thank to the F\# Data contributors on GitHub and other colleagues working on type providers,
including Jomo Fisher, Keith Battocchi and Kenji Takeda.
